rbeast2: BEAUti 2, BEAST2 and Tracer for R

### 2018-02-21 article.tex
Processed feedback, thanks @rsetienne

However, BEAUti cannot be called from a command-line script.
This implies that when the user 
wants to explore the consequences of various settings, this must be done manually.
This is the common workflow when using a few alignments and doing a superficial 
analysis of sensitivity of the reconstructed tree to model settings. 
For exploring many trees (for instance from simulations) and for
more thorough sensitivity analysis, one would like to loop through 
multiple (simulated) alignments, nucleotide substitution models, 
clock models and tree priors. 
One such tool to replace 
BEAUti is \verb;BEASTmasteR; \cite{beastmaster},
which focuses on morphological traits and tip-dating, but also 
supports DNA data.

@@

However, for exploring the output of many BEAST2 runs, 
one would like a script to collect all parameters' ESSes,
parameter traces and posterior phylogenies. 
Examples of R packages to do some of these
are \verb;rBEAST; \cite{rBEAST_ratmann} and 
\verb;RBeast; \cite{RBeast_faria}.

@@

\verb;rbeast2; is the first R package that unifies the
full workflow of working with BEAST2. 
Where \verb;BEASTmasteR; needs dozens for lines to create
a BEAST2 configuration file, with \verb;rbeast2; the same 
can be created by a simple one-liner. Unlike \verb;rBEAST; 
and \verb;RBeast;, \verb;rbeast2; implements all of Tracer's
primary functionality. This makes \verb;rbeast2; a complete 
solution for using BEAST2 from R.

@@

\verb;rbeast2; is written in the R programming language \cite{R}
and enables the full BEAST2 workflow from an R function call,
in a similar way to what BEAUti, DensiTree and Tracer do.

\verb;rbeast2;'s main function is \verb;run_beast2;, which
configures BEAST2, runs it and parses its output. 
\verb;run_beast2; needs at least the name of a 
FASTA file containing a DNA alignment. 
The default settings for the other arguments of \verb;run_beast2; 
are identical to BEAUti's and BEAST2's default settings.
Per alignment, a site model, clock model and tree prior can be chosen.
Multiple alignments can be used, each with its own (unlinked) site model, 
clock model and tree prior.

@@

Currently, \verb;rbeast2; does not replace all functionality in BEAUti,
as it does not supply 3 out of 7 tree priors, nor does it support RNA
alignments or use of morphological data. The many plug-ins of BEAUti
are not supported by \verb;rbeast2;. \verb;rbeast2; does not support
all command-line arguments of BEAST2, does not supply
the more specialized Tracer analysis options, nor is it as feature-rich
in plotting options as DensiTree. Up until now, the \verb;rbeast2; features 
implemented are those requested by users. Further extension of \verb;rbeast2; 
will have to come from future user requests.

@@

BEAUti, and likewise \verb;rbeast2;, needs at least a FASTA filename
to produce a BEAST2 configuration file. 
In BEAUti, this is achieved by loading a FASTA file, 
then saving an output file using a common
save file dialog. After this, BEAST2 needs to be invoked
on the created configuration file to create multiple files
storing the posterior created by BEAST2. Finally, these output
files must be parsed by either Tracer of DensiTree.
In \verb;rbeast2;, the same is achieved by:

@@

This code will create a (temporary) BEAST2 configuration file,
from the FASTA file with name \verb;anthus_aco.fas;, 
using the same default settings as BEAUti, which are, 
among others, a Jukes-Cantor site model, a strict clock, and a Yule birth tree prior.
\verb;rbeast2; will then execute BEAST2 using that file, and
parses the output. The returned data structure, named \verb;out;, 
is a list of parameter estimates (called \verb;estimates;), posterior 
phylogenies (called \verb;anthus_aco_trees;)
and MCMC operator performance (\verb;operators;).

An example of using a different site model, clock model 
and tree prior is:

\begin{lstlisting}[language=R, floatplacement=H]
out <- run_beast2(
  "anthus_aco.fas",
  site_models = hky_site_model(),
  clock_models = rln_clock_model(),
  tree_priors = bd_tree_prior()
)
\end{lstlisting}

This code uses an HKY site model, a relaxed log-normal clock model and a 
birth-death tree prior, each with their default settings in BEAUti.
Table \ref{tab:functions} shows an overview of all functions to 
create site models, clock models and tree priors.

Note that the arguments' names \verb;site_models;, \verb;clock_models; 
and \verb;tree_priors; are plural, as each of these
can be (a list of) one or more elements. Each of these arguments must 
have the same number of elements, so that each alignment has its
own site model, clock model and tree prior.

An example of two alignments, each with its own site model, is:

@@

\verb;rbeast2; also uses the same default prior distributions as BEAUti 
for each of the site models, clock models and tree priors. 
For example, by default, a Yule tree prior assumes the birth rate 
follows a uniform distribution, 
from minus infinity to plus infinity. 
This assumption implies that negative and positive birth rates are just as likely, 
where a negative birth rate is biologically impossible (note that 
in practice, this usually works out just fine).
One may prefer an exponential distribution instead, 
as this would assume only positive birth rates, 
and makes high birth rates unlikely.

To do this in \verb;rbeast2;:

@@

\begin{lstlisting}[language=R, floatplacement=H]
out <- run_beast2(
  get_path("anthus_aco.fas"),
  tree_priors = create_yule_tree_prior(
    birth_rate_distr = create_exp_distr(
      mean = create_mean_param(value = 1.0)
    )    
  )
)
\end{lstlisting}

Our initial motivation to create \verb;rbeast2; 
was that we wanted to fix the crown age of a phylogeny.
BEAUti assumes that a phylogeny has a crown age that needs to be jointly-estimated
with the phylogeny and other parameters. It does not allow for fixing
the crown age. Without \verb;rbeast2;, one needs to manually edit the BEAST2 
configuration file, which is tedious and prone to errors. 
Fixing the crown ages is especially useful for theoretical experiments,
as this allows for one less source of variation.

This is how to specify a fixed crown age with \verb;rbeast2;:

@@

\verb;rbeast2; allows for the same functionality as Tracer.
Tracer works on the values of the parameter estimates sampled
in the BEAST2 run. This is called the "trace" (hence the name).
The start of the trace is usually discarded, as an MCMC 
algorithm (such as used by BEAST2) first has to converge to
its equilibrium. The start of the trace, called the "burn-in", 
will be removed, as its parameter estimates are not 
representative. By default, Tracer discards the first 10\% of all 
the parameter estimates.

To remove the burn-in from all parameter estimates in \verb;rbeast2;,
the following code can be used:

@@

Tracer shows the ESSes of each posterior's variables.
These ESSes are important to determine the strength of the
inference. As a rule of thumb, an ESS of 200 is acceptable 
for any parameter estimate.

To calculate the effective sample sizes (of all estimated variables) in \verb;rbeast2;:

@@

\verb;rbeast2; uses the Travis CI \cite{travis} 
continuous integration service, which is known to significantly 
increase the the number of bugs exposed \cite{vasilescu2015} and increases
the speed at which new features are added \cite{vasilescu2015}.
\verb;rbeast2; has a 100\% code coverage, which correlates with code quality \cite{horgan1994,del1995correlation}. 
\verb;rbeast2; follows Hadley Wickham's style guide \cite{style_guide}, 
which improves software quality \cite{fang2001}.
\verb;rbeast2; dependends on multiple packages, which are 
\verb;APE; \cite{APE}, 
\verb;beautier; \cite{beautier},
\verb;beastier; \cite{beastier},
\verb;devtools; \cite{devtools},
\verb;geiger; \cite{GEIGER},
\verb;ggplot2; \cite{ggplot2},
\verb;knitr; \cite{knitr},
\verb;phangorn; \cite{phangorn},
\verb;rmarkdown; \cite{rmarkdown},
\verb;seqinr; \cite{seqinr},
\verb;stringr; \cite{stringr},
\verb;testit; \cite{testit} and 
\verb;tracerer; \cite{tracerer}.

>>>>

However, BEAUti cannot be called from a command-line script.
This implies that when the user 
wants to explore the consequences of various settings, this must be done manually.
This is the common workflow when using a few alignments and doing a superficial 
analysis of sensitivity of the reconstructed tree to model settings. 
For exploring many trees (for instance from simulations) and for
more thorough sensitivity analysis, one would like to loop through 
multiple (simulated) alignments, nucleotide substitution models, 
clock models and tree priors. 
One such tool to replace BEAUti is \verb;BEASTmasteR; \cite{beastmaster},
which focuses on morphological traits and tip-dating, but also 
supports DNA data. \verb;BEASTmasteR;, however, takes hundreds of
lines to set up (next to a Miscrosft Excel spreadsheet), 
resulting in only 7 citations, all before 2017.

@@

However, for exploring the output of many BEAST2 runs, 
one would like a script to collect all parameters' ESSes,
parameter traces and posterior phylogenies.
There is no single package that offers a complete solution,
but examples of R packages that offer a partial solution
are \verb;rBEAST; \cite{rBEAST_ratmann} and 
\verb;RBeast; \cite{RBeast_faria}.

@@

\verb;rbeast2; is the first R package that unifies the
full workflow of working with BEAST2. 
Whilst \verb;BEASTmasteR; needs hundreds of lines to create
a BEAST2 configuration file, with \verb;rbeast2; the same 
can be created by a simple one-liner. Unlike \verb;rBEAST; 
and \verb;RBeast;, \verb;rbeast2; implements all of Tracer's
primary functionality. This makes \verb;rbeast2; a complete 
solution for using BEAST2 from R.

@@

\verb;rbeast2; is written in the R programming language \cite{R}
and enables the full BEAST2 workflow from an R function call,
in a similar way to what BEAUti, DensiTree and Tracer do.
\verb;rbeast2;'s main function is \verb;run_beast2;, which
configures BEAST2, runs it and parses its output. 
\verb;run_beast2; needs at least the name of a 
FASTA file containing a DNA alignment. 
The default settings for the other arguments of \verb;run_beast2; 
are identical to BEAUti's and BEAST2's default settings.
Per alignment, a site model, clock model and tree prior can be chosen.
Multiple alignments can be used, each with its own (unlinked) site model, 
clock model and tree prior.

@@

Currently, \verb;rbeast2; does not replace all functionality in BEAUti,
as it does not provide 3 out of 7 tree priors, nor does it support RNA
alignments or use of morphological data. The many plug-ins of BEAUti
are not yet supported by \verb;rbeast2;. \verb;rbeast2; does not support
all command-line arguments of BEAST2, does not provide
the more specialized Tracer analysis options, nor is it as feature-rich
in plotting options as DensiTree. Up until now, the \verb;rbeast2; features 
implemented are those requested by users. Further extension of \verb;rbeast2; 
will be based on future user requests.

@@

BEAUti, and likewise \verb;rbeast2;, needs at least a FASTA filename
to produce a BEAST2 configuration file. 
In BEAUti, this is achieved by loading a FASTA file, 
then saving an output file using a common
save file dialog. After this, BEAST2 needs to be applied to
the created configuration file. It creates multiple files
storing the posterior. These output
files must be parsed by either Tracer of DensiTree.
In \verb;rbeast2;, all this is achieved by:

@@

This code will create a (temporary) BEAST2 configuration file,
from the FASTA file with name \verb;anthus_aco.fas;, 
using the same default settings as BEAUti, which are, 
among others, a Jukes-Cantor site model, a strict clock, and a Yule birth tree prior.
\verb;rbeast2; will then execute BEAST2 using that file, and
parses the output. The returned data structure, named \verb;out;, 
is a list of parameter estimates (called \verb;estimates;), posterior 
phylogenies (called \verb;anthus_aco_trees;)
and MCMC operator performance (\verb;operators;).
An example of using a different site model, clock model 
and tree prior is:

\begin{lstlisting}[language=R, floatplacement=H]
out <- run_beast2(
  "anthus_aco.fas",
  site_models = create_hky_site_model(),
  clock_models = create_rln_clock_model(),
  tree_priors = create_bd_tree_prior()
)
\end{lstlisting}

This code uses an HKY site model, a relaxed log-normal clock model and a 
birth-death tree prior, each with their default settings in BEAUti.
Table \ref{tab:functions} shows an overview of all functions to 
create site models, clock models and tree priors.
Note that the arguments' names \verb;site_models;, \verb;clock_models; 
and \verb;tree_priors; are plural, as each of these
can be (a list of) one or more elements. Each of these arguments must 
have the same number of elements, so that each alignment has its
own site model, clock model and tree prior. 
An example of two alignments, each with its own site model, is:

@@

\verb;rbeast2; also uses the same default prior distributions as BEAUti 
for each of the site models, clock models and tree priors. 
For example, by default, a Yule tree prior assumes that the birth rate 
follows a uniform distribution, 
from minus infinity to plus infinity. 
This assumption implies that negative and positive birth rates are just as likely, 
where a negative birth rate is biologically impossible (note that 
in practice, this usually works out just fine).
One may prefer an exponential distribution instead, 
as this would assume only positive birth rates, 
and makes high birth rates unlikely.
To do this in \verb;rbeast2;:

@@

\begin{lstlisting}[language=R, floatplacement=H]
out <- run_beast2(
  "anthus_aco.fas",
  tree_priors = create_yule_tree_prior(
    birth_rate_distr = create_exp_distr(
      mean = create_mean_param(value = 1.0)
    )    
  )
)
\end{lstlisting}

Our initial motivation to create \verb;rbeast2; 
was that we wanted to fix the crown age of a phylogeny.
BEAUti assumes that a phylogeny has a crown age that needs to be jointly-estimated
with the phylogeny and other parameters. It does not allow for fixing
the crown age. Without \verb;rbeast2;, one needs to manually edit the BEAST2 
configuration file, which is tedious and prone to errors. 
Fixing the crown ages is especially useful for theoretical experiments,
as this allows for one less source of variation.
This is how to specify a fixed crown age with \verb;rbeast2;:

@@

\verb;rbeast2; allows for the same functionality as Tracer.
Tracer works on the values of the parameter estimates sampled
in the BEAST2 run. This is called the "trace" (hence the name).
The start of the trace is usually discarded, as an MCMC 
algorithm (such as used by BEAST2) first has to converge to
its equilibrium. The start of the trace, called the "burn-in", 
will be removed, as its parameter estimates are not 
representative. By default, Tracer discards the first 10\% of all 
the parameter estimates. 
To remove the burn-in from all parameter estimates in \verb;rbeast2;,
the following code can be used:

@@

Tracer shows the ESSes of each posterior's variables.
These ESSes are important to determine the strength of the
inference. As a rule of thumb, an ESS of 200 is acceptable 
for any parameter estimate.
To calculate the effective sample sizes (of all estimated variables) in \verb;rbeast2;:

@@

\verb;rbeast2; uses the Travis CI \cite{travis} 
continuous integration service, which is known to significantly 
increase the number of bugs exposed \cite{vasilescu2015} and increases
the speed at which new features are added \cite{vasilescu2015}.
\verb;rbeast2; has a 100\% code coverage, which correlates with code quality \cite{horgan1994,del1995correlation}. 
\verb;rbeast2; follows Hadley Wickham's style guide \cite{style_guide}, 
which improves software quality \cite{fang2001}.
\verb;rbeast2; dependends on multiple packages, which are 
\verb;APE; \cite{APE}, 
\verb;beautier; \cite{beautier},
\verb;beastier; \cite{beastier},
\verb;devtools; \cite{devtools},
\verb;geiger; \cite{GEIGER},
\verb;ggplot2; \cite{ggplot2},
\verb;knitr; \cite{knitr},
\verb;phangorn; \cite{phangorn},
\verb;rmarkdown; \cite{rmarkdown},
\verb;seqinr; \cite{seqinr},
\verb;stringr; \cite{stringr},
\verb;testit; \cite{testit} and 
\verb;tracerer; \cite{tracerer}.
